### 2022-03-28 paper/localpsf.tex
reorganizing intro based on Noemi's suggestions

We present a fast matrix free method for approximating operators that have locally supported non-negative integral kernels. Such operators arise as Schur complements in Schur complements methods for solving partial differential equations (PDEs) \cite{TOBYPAPER}, Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps), covariance operators in spatial statistics, blurring operators in imaging, and Hessians in distributed parameter PDE-constrained optimization and inverse problems.

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain (in applications typically $\gdim=1$, $2$, or $3$) and let $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ be an integral operator of the following form:
\begin{equation}
	\label{eq:kernel_representation}
	(\Aop u)(v) := \int_\Omega \int_\Omega v(y) \Aker(y,x) u(x) dx dy.
\end{equation}
The function $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$ is the integral kernel, $\Aop u \in L^2(\Omega)'$ is the linear functional that results from applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$. We seek to approximate operators $\Aop$ that are accessible \emph{matrix-free}. That is, we have a black-box computational procedure through which we may evaluate the maps
\begin{equation*}
u \mapsto\Aop u \qquad \text{and} \qquad v \mapsto \Aop^Tv
\end{equation*} 
for arbitrary functions $u$ and $v$, but we cannot easily compute kernel entries $\Aker(y,x)$. In applications, evaluating these maps may involve iteratively solving a large linear system, timestepping, or performing some other nontrivial computational procedure. After discretization, $\Aop$ becomes a dense matrix that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means one can compute matrix-vector products with the matrix representation of $\Aop$, but cannot easily compute individual matrix entries.

\begin{figure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch1.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch12.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch19.png}
	\end{subfigure}
	\caption{Batches of normalized impulse responses for the Stokes inverse problem Hessian. Left: 2nd batch. Mid: 13th batch. Right: 19th batch. Brightness scale differs between the subfigures. Black stars indicate the sample point locations at which the point source is applied. Dashed gray ellipses indicate the estimated support ellipsoids. The large circle is the boundary of $\Omega$. Early batches tend to have lots of ellipsoids near the boundary, as compared to later batches. This is because our greedy ellipsoid packing algorithm tries to pick sample points that are far away from sample points in previously chosen batches.}
	\label{fig:batches_intro}
\end{figure}

The output of our method is a hierarchical matrix (H-matrix) approximation, $\mathbf{A}_H$, of a discretization of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank. For more details about H-matrices, see Appendix \ref{app:h_matrix}. While operator actions of $\Aop$ on vectors are typically computationally expensive, fast H-matrix methods permit us to perform matrix-vector products with $\mathbf{A}_H$ cheaply. Furthermore, one may use H-matrix methods to perform useful linear algebra operations involving $\mathbf{A}_H$ that cannot easily be done using $\Aop$, because $\Aop$ is only available matrix-free. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification.

\subsection{Motivation: PDE constrained inverse problems}
\label{sec:PDE_hessian_motivation}

Our motivation for this work is approximation of Hessians in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field, $m$, from noisy observations, 
\begin{equation*}
	y = f(m, u(m)) + \text{noise},
\end{equation*}
which depend on  a state variable $u$. In turn, $u$ depends on $m$ implicitly through the solution of a PDE, which we write generically as follows:
\begin{equation}
	\label{eq:state_pde}
	0 = g(m,u).
\end{equation}
Here $u(m)$ denotes the solution of the PDE \eqref{eq:state_pde} as a function of $m$. In the deterministic approach to inverse problems, one typically finds $m$ as the solution to the minimization problem 
\begin{equation}
	\label{eq:minimization_problem}
	\min_m \quad \frac{1}{2}\|y - f(m,u(m))\|_W^2 + \mathcal{R}(m),
\end{equation}
where $\|\cdot\|_W$ is weighted norm which depends on the noise covariance, and $\mathcal{R}(m)$ is a regularization term. The objective function in optimization problem \eqref{eq:minimization_problem} is defined indirectly via the implicit function theorem. As a result, information about the Hessian of the objective function, $\mathcal{H}$, is only accessible via the action of $\mathcal{H}$ action on vectors. In particular, applying $\mathcal{H}$ to a vector requires solving two PDEs \cite{HESSIANACTION}. More accessible approximations of $\mathcal{H}$ are highly desirable, because they allow for fast solution of \eqref{eq:minimization_problem} via Newton-type methods. More accessible Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because $\mathcal{H}^{-1}$ locally approximates the Bayesian posterior covariance for $m$.

\subsection{Background}

@@

Since we are building an H-matrix approximation of $\Aop$, it is natural to ask why we compute impulse responses and interpolate them. Why not build the H-matrix directly? The reason is that \emph{classical H-matrix construction methods \cite{HMATRIXCONSTRUCTION} require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available through matrix-vector products}. The impulse response interpolation allows fast evaluation of approximate matrix entries needed by the classical H-matrix construction methods. Latest developments and improvements in matrix-free H-matrix construction and approximation are based on the so-called ``peeling process'' \cite{LEXINGPEELINGPROCESS}. These methods have, for instance, been applied to form H-matrix representations of Hessians in PDE constrained inverse problems \cite{ILONAHMATRIX,NOEMIHMATRIX}. One issue with this approach is that, while these matrix-free H-matrix construction methods are asymptotically scalable in theory, in practice the required number of matrix vector products is large\footnote{DO CALCULATION}. Matrix-free H-matrix construction methods are currently a topic of considerable research, so future algorithmic improvements may make matrix-free H-matrix methods based on the peeling process more feasible in the future.

>>>>

We present a fast \emph{matrix free} method for approximating operators, $\Aop$, that have locally supported non-negative integral kernels. Such operators arise as Schur complements in Schur complements methods for solving partial differential equations (PDEs) \cite{TOBYPAPER}, Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps), covariance operators in spatial statistics, blurring operators in imaging, and Hessians in distributed parameter PDE-constrained optimization and inverse problems. By ``matrix free,'' we mean that we have a black-box computational procedure through which we may apply $\Aop$ and $\Aop^T$ to arbitrary functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad v \mapsto \Aop^Tv,
\end{equation} 
but we do not have any further access to $\Aop$. In particular, we cannot easily access entries of $\Aop$'s integral kernel. In applications, evaluating the maps in \eqref{eq:eval_maps} may involve iteratively solving a large linear system, timestepping, or performing some other nontrivial computational procedure. After discretization, $\Aop$ becomes a dense matrix that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means one can compute matrix-vector products with the matrix representation of $\Aop$, but cannot easily compute individual matrix entries.

%We seek to approximate operators $\Aop$ that are accessible \emph{matrix-free}. That is, we have a black-box computational procedure through which we may evaluate the maps
%\begin{equation*}
%u \mapsto\Aop u \qquad \text{and} \qquad v \mapsto \Aop^Tv
%\end{equation*} 
%for arbitrary functions $u$ and $v$, but we cannot easily compute kernel entries $\Aker(y,x)$. In applications, evaluating these maps may involve iteratively solving a large linear system, timestepping, or performing some other nontrivial computational procedure. After discretization, $\Aop$ becomes a dense matrix that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means one can compute matrix-vector products with the matrix representation of $\Aop$, but cannot easily compute individual matrix entries.

The output of our method is a hierarchical matrix (H-matrix). H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank. For more details about H-matrices, see Appendix \ref{app:h_matrix}. H-matrix methods permit us to perform matrix-vector products extremely cheaply. Furthermore, one may use H-matrix methods to perform useful linear algebra operations involving the H-matrix that cannot easily be done using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification.

Since we are building an H-matrix, it is natural to ask why we do not build the H-matrix directly. The reason is that \emph{classical H-matrix construction methods \cite{HMATRIXCONSTRUCTION} require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available through matrix-vector products}. The basic idea behind our method is constructing the H-matrix via impulse response interpolation. The impulse response, $\phi_x$, associated with a point $x$ is the function that results from applying $\Aop$ to a delta distribution (point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with collections of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate these impulse responses to evaluate arbitrary entries of the operator's integral kernel. This allows fast evaluation of approximate matrix entries needed by classical H-matrix construction methods.

\begin{figure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch1.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch12.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch19.png}
	\end{subfigure}
	\caption{Batches of normalized impulse responses for the Stokes inverse problem Hessian. Left: 2nd batch. Mid: 13th batch. Right: 19th batch. Brightness scale differs between the subfigures. Black stars indicate the sample point locations at which the point source is applied. Dashed gray ellipses indicate the estimated support ellipsoids. The large circle is the boundary of the domain. 
%		Early batches tend to have lots of ellipsoids near the boundary, as compared to later batches. 
%		This is because our greedy ellipsoid packing algorithm tries to pick sample points that are far away from sample points in previously chosen batches.
	}
	\label{fig:batches_intro}
\end{figure}

%\subsection{Motivation: PDE constrained inverse problems}
%\label{sec:PDE_hessian_motivation}

Our motivation for this work is approximation of Hessians in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field, $m$, from noisy observations, 
\begin{equation*}
y = f(m, u(m)) + \text{noise},
\end{equation*}
which depend on  a state variable $u$. In turn, $u$ depends on $m$ implicitly through the solution of a PDE, which we write generically as follows:
\begin{equation}
\label{eq:state_pde}
0 = g(m,u).
\end{equation}
Here $u(m)$ denotes the solution of the PDE \eqref{eq:state_pde} as a function of $m$. In the deterministic approach to inverse problems, one typically finds $m$ as the solution to the minimization problem 
\begin{equation}
\label{eq:minimization_problem}
\min_m \quad \frac{1}{2}\|y - f(m,u(m))\|_W^2 + \mathcal{R}(m),
\end{equation}
where $\|\cdot\|_W$ is weighted norm which depends on the noise covariance, and $\mathcal{R}(m)$ is a regularization term. The objective function in optimization problem \eqref{eq:minimization_problem} is defined indirectly via the implicit function theorem. As a result, information about the Hessian of the objective function is only accessible via the action of the Hessian on vectors. In particular, applying the Hessian to a vector requires solving two PDEs \cite{HESSIANACTION}. More accessible approximations of the Hessian are highly desirable, because they allow for fast solution of \eqref{eq:minimization_problem} via Newton-type methods. More accessible Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because inverse of the Hessian locally approximates the Bayesian posterior covariance for $m$.

%\subsection{Background}

@@

Latest developments and improvements in matrix-free H-matrix construction and approximation are based on the so-called ``peeling process'' \cite{LEXINGPEELINGPROCESS}. These methods have, for instance, been applied to form H-matrix representations of Hessians in PDE constrained inverse problems \cite{ILONAHMATRIX,NOEMIHMATRIX}. One issue with this approach is that, while these matrix-free H-matrix construction methods are asymptotically scalable in theory, in practice the required number of matrix vector products is large\footnote{DO CALCULATION}. Matrix-free H-matrix construction methods are currently a topic of considerable research, so future algorithmic improvements may make matrix-free H-matrix methods based on the peeling process more feasible in the future.

\section{Setting}

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain (in applications typically $\gdim=1$, $2$, or $3$) and let $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ be an integral operator of the following form:
\begin{equation}
	\label{eq:kernel_representation}
	(\Aop u)(v) := \int_\Omega \int_\Omega v(y) \Aker(y,x) u(x) dx dy.
\end{equation}
The function $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$ is the integral kernel, $\Aop u \in L^2(\Omega)'$ is the linear functional that results from applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$.

%We seek to approximate operators $\Aop$ that are accessible \emph{matrix-free}. That is, we have a black-box computational procedure through which we may evaluate the maps
%\begin{equation*}
%u \mapsto\Aop u \qquad \text{and} \qquad v \mapsto \Aop^Tv
%\end{equation*} 
%for arbitrary functions $u$ and $v$, but we cannot easily compute kernel entries $\Aker(y,x)$. In applications, evaluating these maps may involve iteratively solving a large linear system, timestepping, or performing some other nontrivial computational procedure. After discretization, $\Aop$ becomes a dense matrix that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means one can compute matrix-vector products with the matrix representation of $\Aop$, but cannot easily compute individual matrix entries.
